---
title: "The Code Book"
output: 
  html_document: 
    keep_md: yes
---

##Introduction

The purpose of cleaning data is to create a "tidy data set"" from the raw data files, structuring the data set to allow for further analysis. Cleaning and preparing the data is an iterative process which facilitates a deeper understanding of the data. This Code Book contains the description of the "tidy_data.txt", including  a description of the study, the variables and the summary of analysis performed on the raw data files.


##Tidy Data set Description
 
The Tidy Data text file consists of the original raw data files merged and analyzed to create a final tidy data set which contains 68 unique variables and 180 different observations of those variables. A data set that meets the following criteria is considered to be “tidy”.

The tidy data definition:

1. Each variable forms a column
2. Each observation forms a row
3. Each data set contains information on only one observational unit of analysis

##Study Design

One of the most exciting areas in all of data science right now is wearable computing. Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users.

The Tidy Dataset describes the participants of an experiment who performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. The subjects where a group of 30 volunteers between the ages of 19 and 48 years old. The Raw Data captured for each subject where 3-axial linear accelerations and 3-axial angular velocities at a constant rate of 50Hz using the embedded accelerometer and gyroscope. A full description of the Raw Data is available at the site where the data was obtained: 

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones 

##The Code book Variables

###List of Variables from the Tidy Data
```{r}
tidy_data <- read.table("tidy_dataset.txt", header = TRUE) 
```

The following are the full list variables the Tidy Dataset:

```{r}
names(tidy_data)
```
###Description of Variables

The subject and activity labels described each individual who participated in the experiment and which activities they performed. Each of the 30 individual subjects completed 6 different activities giving a total of 180 observations for each variable.

The variables are normalized and bounded within [-1,1].
The units used for the accelerations (total and body) are 'g's (gravity of earth -> 9.80665 m/seg2).
The gyroscope units are rad/seg.

The variables selected for this data set come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

###The set of variables that were included are: 
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

- tBodyAcc-XYZ
- tGravityAcc-XYZ
- tBodyAccJerk-XYZ
- tBodyGyro-XYZ
- tBodyGyroJerk-XYZ
- tBodyAccMag
- tGravityAccMag
- tBodyAccJerkMag
- tBodyGyroMag
- tBodyGyroJerkMag
- fBodyAcc-XYZ
- fBodyAccJerk-XYZ
- fBodyGyro-XYZ
- fBodyAccMag
- fBodyAccJerkMag
- fBodyGyroMag
- fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

- mean: Mean value
- std: Standard deviation

###Data Analysis Summary

The purpose of the analysis is to create an independent tidy data set with the average of each variable for each activity and each subject. 

The UCI HAR data set is unzipped and the test and train files for X, Y, and subject, as well as the features and activity files,  are read into the working directory. For both the test and train files, the activity labels are joined with the activities performed by the subjects. After removing the first column in the resulting table, the activities are renamed from numerals using the descriptive activity labels. The columns of the X data are renamed using the features.txt file. 

The test and train files are then column bound to create a merged test and merged train files which include 561 variables and 2947 and 7352 observations. These test and train tables are then row bound together to create a final data set of 561 variables and 10299 observations. 

From this data set, the variables containing the data for mean and standard deviation are selected. The data set is grouped by Subject and then by activity and the average for each of these variables are recorded. The resulting data set containes 68 variables and 180 observations, which come from the 30 subjects each performing 6 activities. 

Finally, the variable names are expanded and simplified for the user's ease of understanding. This tidy data set is then written to a text file and read in and openned for viewing.
